---
title: "Fitting traits to data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting traits to data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
```

Fitting an optimality model to gas-exchange data used to mean choosing between two
bad options. Either accept a finite-difference gradient, which costs two full
model solves per parameter per observation and returns a number whose accuracy
depends on a step size you have to guess, or give up on gradients and use a
derivative-free optimiser that scales badly past a handful of parameters.

This package removes that choice. Two things make the difference.

**The model itself is fast.** It is a set of header-only C++ routines with no R in
the solve path, and one operating point costs about **3 µs**. A likelihood over a
thousand observations is therefore milliseconds of model, which means the fit is
no longer bounded by how long the biology takes to evaluate.

**And it differentiates itself, exactly.** Forward-mode automatic differentiation
supplies the derivatives of the photosynthesis and cost algebra, and the implicit
function theorem carries them through the two nested solvers, so
`leaf_gradient()` returns the exact derivative of every solved output with respect
to every parameter — with no step size to choose and no truncation error. The
composite is assembled in C++ and vectorised over observations, at **10.6 µs per
observation** for four differentiated parameters.

Put together, those change what is affordable. A finite-difference fit costs two
solves for every parameter *the optimiser is moving*. The exact route costs a term
in the number of parameters *the model has*, and those two counts are only equal
in the simplest possible design. The moment you pool, add a hierarchy, or share a
parameter across species — which is to say, the moment the fit becomes
interesting — the first count grows and the second does not:

| a fit with | finite differences | exact gradient |
|---|---:|---:|
| 3 parameters, 72 observations | fast enough not to care | fast enough not to care |
| 40 parameters over 4 model parameters, 1,327 observations | 2.3 s | **0.67 s** |
| 57 parameters over the same 4 | 3.3 s | **0.67 s**, unchanged |

That last row is the point. Adding seventeen more fitted parameters costs the
exact route nothing at all, because the model still has four.

This vignette works through such a fit from the beginning, and is honest about the
three things that make it harder than it looks.

## Simulated data, so the answer is known

Recovery error is the only fit diagnostic that distinguishes "the optimiser
converged" from "the model works". Real data cannot supply it, so start here.

```{r}
set.seed(sum(utf8ToInt("phylloptim-fitting")))

truth <- leaf_traits(vcmax_25 = 85, jmax_25 = 140, TF24_cost_scale = 40)

drivers <- expand.grid(
  psi_soil = seq(0.4, 3.2, length.out = 12),
  PPFD = c(400, 900, 1500),
  atm_vpd = c(1.0, 2.5)
)

obs <- leaf_solve(psi_soil = drivers$psi_soil, PPFD = drivers$PPFD,
                  atm_vpd = drivers$atm_vpd, traits = truth)

sigma <- c(A = 0.4, gc = 0.0015, psi_stem = 0.08)
obs$A_obs <- obs$A + rnorm(nrow(obs), 0, sigma[["A"]])
obs$gc_obs <- obs$gc + rnorm(nrow(obs), 0, sigma[["gc"]])
obs$psi_obs <- obs$psi_stem + rnorm(nrow(obs), 0, sigma[["psi_stem"]])

nrow(obs)
```

Three responses, because they constrain different parts of the model: `psi_stem`
pins the hydraulic path, `A` and `gc` pin the carbon side. Fitting `A` alone
leaves the two trading off against each other.

## The objective, and the loop that makes it affordable

```{r}
FIT <- c("vcmax_25", "jmax_25", "TF24_cost_scale")

predict_at <- function(theta, l) {
  set_traits(l, do.call(leaf_traits, as.list(exp(theta))))
  out <- vapply(seq_len(nrow(obs)), function(i) {
    set_drivers(l, psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                atm_vpd = obs$atm_vpd[i])
    l$find_root_collar_psi()
    c(l$assim_colimited_, l$stom_cond_CO2_, l$opt_psi_stem_)
  }, numeric(3))
  list(A = out[1, ], gc = out[2, ], psi = out[3, ])
}

ssq <- function(theta, l) {
  p <- predict_at(theta, l)
  sum(((obs$A_obs - p$A) / sigma[["A"]])^2) +
    sum(((obs$gc_obs - p$gc) / sigma[["gc"]])^2) +
    sum(((obs$psi_obs - p$psi) / sigma[["psi_stem"]])^2)
}
```

Two things about that loop are worth getting right, because both are easy to get
wrong and neither fails loudly.

**Build the model once, outside the loop.** `set_traits()` replaces the traits on
an existing object. Constructing a fresh one costs about 45 solves, against 3.4 to
re-trait, so building inside the loop throws away roughly 40 solves per objective
evaluation. In a fit, that dominates everything else.

**Do not write `l$vcmax_25 <- x` instead.** It compiles, and it is wrong. Two
pieces of derived state go on describing the old trait afterwards, and no
re-driving repairs either. The first is the pre-integrated vulnerability splines,
one owned by `stem_P50`/`stem_c` and one by `root_P50`/`root_c`. The second is the
solved operating point, which no code path rewrites, so it reads as though it
belonged to the new traits.

A bare write also bypasses the positive-magnitude checks on the two `P50`
parameters, and it cannot reach `stem_b` or `psi_crit` at all, since those are
derived and read-only. `set_traits()` clears all of it, and is the only route
bound to R.

The temperature-dependent `vcmax_`, `jmax_` and `R_d_` are not part of this
problem. Their cache is keyed on every input of the temperature block, including
`vcmax_25` and `jmax_25`, so re-driving after a bare trait write does recompute
them. It is the splines and the operating point that no re-driving repairs.

```{r}
l <- leaf_model(truth)
start <- log(c(vcmax_25 = 60, jmax_25 = 100, TF24_cost_scale = 15))
fit <- optim(start, ssq, l = l, method = "Nelder-Mead",
             control = list(maxit = 2000, reltol = 1e-10))

data.frame(
  parameter = FIT,
  truth = unlist(truth[FIT]),
  recovered = round(exp(fit$par), 2),
  `pct error` = round(100 * (exp(fit$par) - unlist(truth[FIT])) /
                        unlist(truth[FIT]), 2),
  check.names = FALSE
)
```

## Using the gradients

`leaf_gradient()` gives exact derivatives of each output, so the objective's
gradient follows by the chain rule, with no differencing of the solve.

This fit names three of the five columns, which is the normal case here. `A`, `gc`
and `psi_stem` are measured; `collar` and `profit` are not. `profit` is in the
matrix for a demographic consumer rather than a calibration one, since `plant`
bills carbon from it, so a fit against gas exchange simply ignores that column. It
costs nothing to leave there, because the whole row is read off one solve.

```{r}
ssq_grad <- function(theta, l) {
  tr <- do.call(leaf_traits, as.list(exp(theta)))
  g <- vapply(seq_len(nrow(obs)), function(i) {
    gi <- leaf_gradient(psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                        atm_vpd = obs$atm_vpd[i], traits = tr, pars = FIT)
    v <- gi$value
    r <- c((v[["A"]] - obs$A_obs[i]) / sigma[["A"]]^2,
           (v[["gc"]] - obs$gc_obs[i]) / sigma[["gc"]]^2,
           (v[["psi_stem"]] - obs$psi_obs[i]) / sigma[["psi_stem"]]^2)
    2 * (gi$gradient[FIT, "A"] * r[1] +
           gi$gradient[FIT, "gc"] * r[2] +
           gi$gradient[FIT, "psi_stem"] * r[3])
  }, numeric(length(FIT)))
  rowSums(g) * exp(theta)   # chain rule for the log parameterisation
}
```

Note `pars = FIT`. Requesting all sixteen parameters here would compute twelve
gradients per observation that the optimiser never reads.

**Bound the parameters, or the first step overflows.** The objective is scaled
by `1/sigma^2` and `sigma_gc` is 0.0015, so the gradient has magnitude ~2e4 in log
units. Plain `method = "BFGS"` starts from an identity Hessian approximation, which
makes its first trial point `theta - g`, and `exp()` of that is `Inf`, so the
objective errors out before the line search gets a chance to shrink the step. This
is not a pathology of the model; it is what an unscaled least-squares objective
does to any quasi-Newton method started cold. Bounds are the one-line fix, and are
what you would want on a real fit anyway.

```{r}
lower <- log(c(vcmax_25 = 10, jmax_25 = 20, TF24_cost_scale = 2))
upper <- log(c(vcmax_25 = 400, jmax_25 = 700, TF24_cost_scale = 400))

fit_g <- optim(start, ssq, gr = ssq_grad, l = l, method = "L-BFGS-B",
               lower = lower, upper = upper, control = list(maxit = 500))

fit_n <- optim(start, ssq, l = l, method = "L-BFGS-B",
               lower = lower, upper = upper, control = list(maxit = 500))

data.frame(
  method = c("Nelder-Mead (no gradient)", "L-BFGS-B (numerical gradient)",
             "L-BFGS-B (exact gradient)"),
  objective = round(c(fit$value, fit_n$value, fit_g$value), 2),
  `objective evals` = c(fit$counts[["function"]], fit_n$counts[["function"]],
                        fit_g$counts[["function"]]),
  `gradient evals` = c(0, fit_n$counts[["gradient"]], fit_g$counts[["gradient"]]),
  `honest total` = c(fit$counts[["function"]],
                     fit_n$counts[["function"]] +
                       2 * length(FIT) * fit_n$counts[["gradient"]],
                     fit_g$counts[["function"]]),
  check.names = FALSE
)
```

All three find the same optimum. The exact gradient cuts the solve count roughly
five-fold against Nelder-Mead, and that is the whole of its advantage here.

**Counting evaluations fairly is harder than it looks.** `optim` reports
function and gradient calls separately, and when it differences a gradient
*numerically* it does not count those calls at all. BFGS and L-BFGS-B use central
differences, so each numerical gradient silently costs `2 × npar` extra objective
evaluations, which the `honest total` column above adds back. Comparing a
derivative-free method's reported count against a gradient method's reported count
without that correction flatters the gradient method by a factor of `2 × npar`.

## Faster or slower depends on which parameter count you mean

Evaluation counts are not wall clock, and here the two disagree: at three
parameters the exact gradient **loses**. That is a real result, and it is also the
most misread one in this package, because it does not generalise the way it looks
like it should. Both arms scale with the number of parameters, and they scale with
**different** parameter counts.

Measure both arms in one process, interleaved, over a growing request:

```{r}
#| label: scaling
tm <- function(f) {
  t0 <- Sys.time(); f(); 1000 * as.numeric(difftime(Sys.time(), t0, units = "secs"))
}

CAND <- c("vcmax_25", "jmax_25", "TF24_cost_scale", "TF24_beta2", "a",
          "curv_fact_elec_trans", "curv_fact_colim",
          "stem_P50", "stem_c", "root_P50", "root_c")

exact_n <- function(pars) {
  vapply(seq_len(nrow(obs)), function(i) {
    leaf_gradient(psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                  atm_vpd = obs$atm_vpd[i], traits = truth,
                  pars = pars)$gradient[pars, "A"]
  }, numeric(length(pars)))
}
ssq_at <- function(theta, l, nm) {
  tr <- truth; tr[nm] <- as.list(exp(theta))
  set_traits(l, do.call(leaf_traits, tr))
  out <- vapply(seq_len(nrow(obs)), function(i) {
    set_drivers(l, psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                atm_vpd = obs$atm_vpd[i])
    l$find_root_collar_psi()
    c(l$assim_colimited_, l$stom_cond_CO2_, l$opt_psi_stem_)
  }, numeric(3))
  sum(((obs$A_obs - out[1, ]) / sigma[["A"]])^2) +
    sum(((obs$gc_obs - out[2, ]) / sigma[["gc"]])^2) +
    sum(((obs$psi_obs - out[3, ]) / sigma[["psi_stem"]])^2)
}
fd_n <- function(pars, l) {
  th <- log(unlist(truth[pars]))
  for (j in seq_along(pars)) {
    a <- th; b <- th; a[j] <- a[j] + 1e-5; b[j] <- b[j] - 1e-5
    ssq_at(a, l, pars); ssq_at(b, l, pars)
  }
}

for (k in 1:3) { invisible(exact_n(CAND[1:2])); fd_n(CAND[1:2], l) }  # warm both

scaling <- do.call(rbind, lapply(c(1, 3, 5, 8, 11), function(n) {
  p <- CAND[seq_len(n)]
  r <- replicate(3, c(tm(function() exact_n(p)), tm(function() fd_n(p, l))))
  data.frame(n_par = n, exact_ms = round(min(r[1, ]), 1),
             fd_ms = round(min(r[2, ]), 1))
}))
scaling$ratio <- round(scaling$fd_ms / scaling$exact_ms, 2)
scaling
```

```{r}
#| label: scaling-slopes
rbind(
  exact = round(coef(lm(exact_ms ~ n_par, scaling)), 2),
  fd = round(coef(lm(fd_ms ~ n_par, scaling)), 2)
)
```

Both are linear, and the two coefficients are the whole story:

- **the finite difference is pure slope.** Its intercept is about 0 and its slope is
  `2 ×` one objective evaluation, because that is literally what it does;
- **the exact gradient has a large intercept and a slope too.** The intercept is
  the per-observation `Leaf` construction and solve, paid once whatever you ask
  for. The slope is `2` evaluations per parameter per observation.

Here the exact route's *slope* is the larger of the two, so **growing the parameter
count in this design never makes it win.** The ratio improves but does not reach 1.
Each extra parameter costs it two more R calls per observation, and from R a call
costs more than the C++ work it wraps.

### What reverses it is more *fitted* parameters per *model* parameter

The two counts are the same thing only because this vignette fits traits directly.
They come apart as soon as a parameterisation sits between the fit and the model:

$$T_\text{FD} = 2\,P_\text{fit}\,T_\text{obj}
\qquad
T_\text{exact} = N\,(t_\text{construct} + t_\text{solve}) + 2\,P_\text{model}\,N\,t_\text{eval}$$

**`leaf_gradient()` is asked for `P_model` parameters; `optim` differences
`P_fit` of them.** A hierarchical model is the standard case where those differ: a
non-centred population mean, SD and per-species deviation are three fitted
parameters that reach the leaf as *one*.

A companion calibration study is that case, and it is worth quoting because it
comes out the other way round. 1,327
observations, 16 species, **40 fitted parameters that map onto 4 model
parameters**:

| | |
|---|---|
| exact gradient, 40 fitted parameters | **0.68 s** |
| exact gradient, 57 fitted parameters | **0.67 s**, unchanged, because `P_model` is still 4 |
| central differences, 40 fitted parameters | **2.12 s** |
| the resulting fit | same optimum, **3.4× less wall clock** |

**The cost model is not fitted to that study. It predicts it.** Take the two
coefficients measured above, divide by this vignette's 72 observations to get
per-observation costs, and substitute that study's design:

```{r}
#| label: predict-other-study
co <- coef(lm(exact_ms ~ n_par, scaling)) / nrow(obs)   # ms per observation
LC <- list(N = 1327, P_model = 4, T_obj_ms = 26,        # measured there
           exact_ms = 679, break_even = 13.1)

predicted <- LC$N * (co[[1]] + co[[2]] * LC$P_model)
data.frame(
  quantity = c("exact gradient (ms)", "break-even fitted parameters"),
  predicted = round(c(predicted, predicted / (2 * LC$T_obj_ms)), 1),
  measured_there = c(LC$exact_ms, LC$break_even)
)
```

Two designs differing 18-fold in size, one set of coefficients, agreeing to a few
percent. That is the evidence that the scaling law is the mechanism and not a
coincidence of either design.

**So the rule to take away is not "AD is slower here".** It is:

> Count how many *distinct model parameters* your gradient needs, not how many
> parameters your optimiser is moving. Differencing costs the latter; the composite
> costs the former.

If those numbers are equal, difference the objective. If your parameterisation
makes the first much smaller than the second, which pooling, hierarchy and any
shared or derived parameter all do, use the gradient.

### Where the intercept goes

The remaining measurement, since the intercept is the exact route's whole
disadvantage in this design:

```{r}
fd_gradient <- function() {
  for (j in seq_along(start)) {
    a <- start; b <- start; a[j] <- a[j] + 1e-5; b[j] <- b[j] - 1e-5
    ssq(a, l); ssq(b, l)
  }
}
construct_only <- function() for (i in seq_len(nrow(obs))) leaf_model(truth)

# Warm up all three arms first, then interleave. A cold leaf_model() measured
# three times its steady-state cost, which is enough to turn "a third" into
# "nine tenths" and point the conclusion at the wrong fix.
for (k in 1:3) { ssq_grad(start, l); fd_gradient(); construct_only() }

reps <- replicate(5, c(tm(function() ssq_grad(start, l)),
                       tm(fd_gradient),
                       tm(construct_only)))
data.frame(
  route = c("exact gradient (72 x leaf_gradient)",
            "central differences (2 x npar objective evals)",
            "72 x leaf_model() alone"),
  ms = round(apply(reps, 1, min), 1)
)
```

The third row is most of the intercept. `leaf_gradient()` constructs a fresh model
on every call, and there is no argument for handing it one you already have, so a
gradient over `n` observations pays `n` constructions.

The rest is the R call boundary itself. The composite replaces one solve with a
handful of much cheaper evaluations, but each of those is an R call into C++, and an
R call costs more than the C++ work it wraps. The theorem does less arithmetic and
more crossing. That is also why the exact route has a slope in `P_model` at all,
rather than the near-free one the C++ benchmarks show.

Four things follow:

- **Do not choose between these on the evaluation counts alone**, and do not choose
  on this vignette's verdict either. Put your own `N`, `P_fit` and `P_model` into the
  cost model above, or time it on your own design.
- **`P_fit` = `P_model` is the worst case for the composite**, and it is the case a
  vignette naturally demonstrates because fitting traits directly is the simplest
  thing to show. Real fits are usually not that case.
- **Reuse the model object.** `leaf_gradient(x = )` takes a `Leaf` you already
  have, which removes the construction cost from the intercept. In
  that study, model construction was 29% of the exact gradient. It does not make the
  composite the faster arm in *this* design, because here the problem is the slope
  rather than the intercept.
- **Use the batch entry point, which removes the slope.**
  `leaf_gradient_batch()` composes the theorem in C++ and loops over observations
  there, so the boundary is crossed once per call instead of 112 times per
  observation: **10.6 µs per observation against 363**, at `length(pars)` = 4.

Read the table above as what the R boundary costs, not as the fastest this can
go. The cost *model* is the thing to reason with, and `P_fit` against `P_model` is
what decides which arm wins; the batch route changes only the constant in front of
the exact arm. The same fit through `leaf_batch()` and `leaf_gradient_batch()`:

```r
b <- leaf_batch(psi_soil = obs$psi_soil, PPFD = obs$PPFD)   # once, not per draw
g <- leaf_gradient_batch(b, traits, pars = FIT)
```

with the likelihood and your parameterisation Jacobian still in R, vectorised over
observations, which is where they belong, and is why the `P_fit > P_model`
structure is what the exact route exploits.

## Fitting a different collection of parameters

Nothing above is specific to the parameters it fits. `pars` is an argument, so a
second study wanting a different collection changes that argument and nothing else,
the same prepared `leaf_batch()` serves any of them, because a batch holds the
*drivers*, and which parameters you differentiate is not a property of the data.

```{r}
#| label: collections
b <- leaf_batch(psi_soil = seq(0.6, 3.0, length.out = 40), PPFD = 900,
                atm_vpd = 2)

studies <- list(
  "demand only"      = c("vcmax_25", "jmax_25"),
  "the cost model"   = c("TF24_cost_scale", "TF24_beta2"),
  "hydraulic supply" = c("leaf_specific_conductance_max", "stem_P50", "stem_c"),
  "adds both curves" = c("vcmax_25", "stem_P50", "root_P50")
)

do.call(rbind, lapply(names(studies), function(nm) {
  p <- studies[[nm]]
  t0 <- Sys.time()
  g <- leaf_gradient_batch(b, pars = p)
  ms <- 1000 * as.numeric(Sys.time() - t0, units = "secs")
  # Which parameters came back with NO signal at all, anywhere?
  dead <- p[colSums(abs(g$gradient[, , "A", drop = FALSE])) == 0]
  data.frame(study = nm, n_pars = length(p), ms = round(ms, 2),
             `ms per par` = round(ms / length(p), 2),
             `zero gradient` = if (length(dead)) paste(dead, collapse = ", ")
                               else "-",
             check.names = FALSE)
}))
```

Three things in that table are worth more than the timings, and none of them is
obvious from the interface.

### Cost per parameter is not uniform

Compare the `ms per par` column across rows: the hydraulic study is several times
the others, and that ratio is the point rather than its exact value, which moves
with the machine. `stem_b` and
`stem_c` own the pre-integrated stem vulnerability spline, so perturbing either
rebuilds it, at 11.9 µs of incomplete gamma functions per side, several times the
cost of a whole solve. `stem_b` escapes most of that by a homogeneity identity
(`fast_stem_curve`, on by default); `stem_c` has no such identity and cannot.

So **a fit that frees a vulnerability curve is in a different cost regime from one
that does not**, and it is the regime the exact gradient helps least with, since the
four hydraulic traits are also the ones whose gradient is *entirely*
argmax-mediated, which is to say the ones you most need it for. Budget for it, and
see `?leaf_gradient`'s `fast_stem_curve`.

### A stem gradient means something different at a boundary

This is the sharpest trap on this surface, and the table above provokes it
deliberately.

`psi_crit` sets the dry end of the *feasible* collar interval and appears in the
profit function nowhere else. It is not a parameter you can ask for, being the
5%-conductivity quantile of the stem curve, so its effect arrives as a term
inside `d/dstem_P50`**, and that term behaves completely differently on the two
sides of a pin. At an interior optimum it is exactly zero, and the gradient is the
curve's own effect scaled by `(ln 2)^{-1/c}`. **At a boundary-crit optimum the bound
is binding**, the term is order 1, and at the default traits it is 29% of the
total. `leaf_gradient()` differences the solve there precisely so that half is not
lost.

A silent zero is a worse failure than a wildly wrong number: it tells an optimiser
the parameter does nothing, and nothing about it looks suspicious. If you fit a
threshold parameter, **check `status` alongside the gradient.** An all-interior
dataset cannot inform it at all, and that is a statement about your design rather
than about the solver.

### Not everything is reachable

`resistance` exists only on the single-potential supply path; ask for it on the
multi-layer path and you get an error naming the path rather than "unknown
parameter". `beta_R_H` and `beta_R_V` left with the root-architecture model
entirely, so a fit that needs them has to difference `root_network_from_carbon()`
itself, which is cheap to perturb (it is homogeneous of degree 1 in each) but still
costs two solves.

### What a new study actually has to write

Not the gradient, but the map from its own sampled parameters onto the model's, and
that map's Jacobian. `pars` names four things; an optimiser moving forty needs a
40 × 4 chain rule, and that is the caller's arithmetic because it encodes the
caller's model. `vignette("phylloptim")` and the companion calibration study
both do this; the latter's `leaf_natural_jacobian()` is the worked example, including
the case worth being careful about: **one sampled parameter that reaches two model
quantities, in opposite directions**. Getting a sign wrong there does not fail. It
converges, somewhere else.

## Where a gradient-based fit goes wrong

The derivation behind `leaf_gradient()` assumes the optimum is interior. In dry
soil it often is not, and the boundary is a genuine kink in trait space rather
than an artefact.

```{r}
status <- vapply(seq_len(nrow(obs)), function(i) {
  leaf_gradient(psi_soil = obs$psi_soil[i], PPFD = obs$PPFD[i],
                atm_vpd = obs$atm_vpd[i], traits = truth,
                pars = "vcmax_25")$status
}, character(1))
table(status)
```

**A third of the design sits on a boundary**, and the fit above still recovered the traits
to a few percent, because `leaf_gradient()` detects those points and differences
the solve there instead of applying a formula that would be wrong by orders of
magnitude. That is the fallback doing its job on a perfectly ordinary drought
design, not an exotic case.

Check this *before* trusting a fit rather than after. `status` is reported
precisely so it can be checked, and a naive finite difference cannot tell you
which points are which. Where that fraction is large, expect a
gradient-based optimiser to behave less well than it does here.

Because a fit that straddles the boundary has three problems:

- **the objective is not differentiable there**, so a gradient method is
  optimising through a corner and may stall or oscillate;
- `leaf_gradient()` falls back to differencing the solve at boundary points, which
  is correct but noisier;
- **the stem traits are special.** The dry bound `psi_crit` is a quantile of
  the stem curve, so it moves with `stem_P50`. But it bounds the feasible range
  rather than entering the profit function, so the composite cannot see that
  movement and returns only the curve's own term. At an interior optimum that is
  the whole answer. At a boundary-crit optimum it is 29% short, and nothing about the
  number looks wrong. The fallback exists for exactly this and is not a degraded
  path.

## Fitting to real data

The same machinery, with two additions worth planning for.

**Not every parameter is identified.** Parameters that enter the model only in
combination cannot be separated by data that sees only the combination. Fitting
them independently produces respectable-looking estimates that mean nothing.
Check with a profile, or a Hessian at the optimum, before reporting anything
per-parameter.

**Some traits are measured.** `stem_P50` *is* a published quantity, and a second
quantile of the same curve fixes `stem_c`, so a species with a measured
vulnerability curve needs no hydraulic calibration at all:

```{r}
shape_from_p50_p88 <- function(P50, P88) {
  log(log(1 / 0.12) / log(2)) / log(P88 / P50)
}
w <- c(stem_P50 = 6.88, stem_c = shape_from_p50_p88(P50 = 6.88, P88 = 9.43))
round(w, 4)

lw <- leaf_model(leaf_traits(stem_P50 = w[["stem_P50"]], stem_c = w[["stem_c"]]))
round(c(PLC_at_P50 = 1 - lw$proportion_of_conductivity(6.88),
        PLC_at_P88 = 1 - lw$proportion_of_conductivity(9.43)), 4)
```

Exactly 0.50 and 0.88, as they must be. Fitting what is already measured wastes
the data's ability to constrain what is not.
