---
title: "Leaf gas exchange from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Leaf gas exchange from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_bw(base_size = 11) +
            theme(panel.grid.minor = element_blank(),
                  strip.background = element_rect(fill = "grey95",
                                                  colour = NA)))
```

## What this model does differently

Most leaf gas-exchange models get stomatal conductance from an empirical
function of assimilation, VPD and CO₂ — Ball-Berry, Medlyn, and their
descendants. This one has no such function. It builds an explicit hydraulic path
from the soil, through the roots and the root collar, up the stem to the leaf,
and then asks: **which stem water potential maximises profit?** Profit is
assimilation minus a hydraulic cost that rises as the xylem loses conductivity.
Conductance is whatever falls out of the answer.

So stomatal closure under drought here is not fitted. It is derived, from two
Weibull vulnerability curves, the soil potential, and the plant's own
conductance.

## One solve

```{r}
leaf_solve(psi_soil = 2.0, PPFD = 900) |>
  as_tibble() |>
  glimpse()
```

Drivers in, operating point out. `psi_soil` is a **positive magnitude in MPa** —
more on that below.

Most of those columns are the usual gas-exchange quantities. Two are less usual:

- **`lambda`** is the marginal cost of water, dA/dE, at the operating point —
  how much carbon the leaf gives up per unit of water it declines to transpire.
  It is the quantity the optimisation equalises.
- **`g1_eff`** re-expresses the solved conductance as the `g1` of the Medlyn
  model, which is a convenient common scale if that is the model you are used
  to. It is computed from the solve rather than supplied, so it varies with
  conditions.

## A drought response

`leaf_solve()` is vectorised over its drivers, so a response curve is one call.

```{r}
drought <- leaf_solve(psi_soil = seq(0.2, 5.5, length.out = 40), PPFD = 900) |>
  as_tibble()

drought |>
  select(psi_soil, A, gc) |>
  pivot_longer(c(A, gc), names_to = "variable") |>
  mutate(variable = recode(variable,
                           A = "assimilation~(mu*mol~m^-2~s^-1)",
                           gc = "stomatal~conductance~(mol~m^-2~s^-1)")) |>
  ggplot(aes(psi_soil, value)) +
  geom_line(linewidth = 0.8, colour = "steelblue4") +
  facet_wrap(~variable, scales = "free_y", labeller = label_parsed) +
  labs(x = "soil water potential (MPa)", y = NULL)
```

Note where both hit exactly zero. That is not a floor imposed by a
minimum-conductance parameter — it is the solve reporting that no operating
point both moves water and stays inside the root's hydraulic limit, so the
answer is shut-down.

## The same sweep on a familiar scale

`g1_eff` across the same drought, over the range where the stomata are open:

```{r, fig.height = 3.4}
drought |>
  filter(gc > 0, is.finite(g1_eff)) |>
  ggplot(aes(psi_soil, g1_eff)) +
  geom_line(linewidth = 0.8, colour = "firebrick") +
  labs(x = "soil water potential (MPa)",
       y = expression(effective~g[1]~(kPa^0.5)))
```

It declines as the soil dries, because it is computed from the solve rather than
being a parameter of it.

## Light and VPD together

Because `leaf_solve()` recycles its drivers elementwise, a factorial design is an
`expand_grid()` and a single call — no loop, and nothing to join back afterwards,
since the drivers come back as columns of the answer.

```{r}
expand_grid(PPFD = seq(50, 1800, length.out = 30),
            atm_vpd = c(0.5, 1, 2, 4)) |>
  with(leaf_solve(psi_soil = 1.0, PPFD = PPFD, atm_vpd = atm_vpd)) |>
  ggplot(aes(PPFD, A, colour = factor(atm_vpd))) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_d(end = 0.9, direction = -1) +
  labs(x = expression(PPFD~(mu*mol~m^-2~s^-1)),
       y = expression(A~(mu*mol~m^-2~s^-1)),
       colour = "VPD (kPa)")
```

The curves saturate at different heights because the hydraulic cost of the
transpiration needed to support a given assimilation rises with VPD.

## Traits and numerical settings are separate

The C++ constructor takes seventeen positional arguments, four of which are
tolerances sitting among the physiology. `leaf_traits()` and `leaf_control()`
split them, so a calibration loop varying traits never has to know which of the
seventeen are not traits.

```{r}
vulnerable <- leaf_traits(stem_b = 2.5, psi_crit = 4.0,
                          root_b = 2.5, root_psi_crit = 4.0)

bind_rows(
  default = leaf_solve(psi_soil = 3.0, PPFD = 900),
  vulnerable = leaf_solve(psi_soil = 3.0, PPFD = 900, traits = vulnerable),
  .id = "traits"
) |>
  select(traits, psi_stem, A, gc, g1_eff)
```

`stem_b`/`stem_c` and `root_b`/`root_c` are **two different vulnerability
curves** — the stem pair drives the hydraulic cost, the root pair drives uptake.
They are named rather than left as an unmarked `b`/`c` because using one where
the other belongs is a mistake that has been made here and reached a manuscript
draft.

Tolerances go the other way, through `leaf_control()`:

```{r}
tibble(
  setting = c("default", "tight"),
  psi_stem = c(
    leaf_solve(psi_soil = 2.0, PPFD = 900)$psi_stem,
    leaf_solve(psi_soil = 2.0, PPFD = 900,
               control = leaf_control(GSS_tol_abs = 1e-5))$psi_stem
  )
)
```

Those are **identical**, and this passage used to say they differed in the fourth
decimal. They did, until the collar solve stopped searching profit and started
solving its own first-order condition instead: the operating point is now
determined to about `1e-12` rather than to `GSS_tol_abs`, which no longer reaches
the production path at all. It survives as the threshold below which the feasible
interval is treated as a single point, and for two single-layer optimisers that
are off this path.

## Water potentials are positive magnitudes

There is one representation throughout: every ψ is a positive number of MPa,
including `psi_soil`, `psi_crit` and the collar potential. Where an equation
needs one potential to oppose another, the minus sign is in the equation.

This is asserted rather than documented, which matters if you are porting a
script that passed signed potentials:

```{r, error = TRUE}
leaf_solve(psi_soil = -2.0, PPFD = 900)
```

An error, not a quiet `abs()`. A model that accepted both conventions would let
that script return a plausible wrong number instead.

## A bare leaf: one soil potential, one resistance

Everything above went through the multi-layer root network, which is what plant
uses. At one layer with no vulnerability-weighted horizontal term, that network
is a single soil-to-collar resistance — and note that it is driven through **the
same `root_network` argument**, because the resistances are a per-call driver
either way:

```{r}
leaf_solve(psi_soil = 1.5, PPFD = 900,
           supply = leaf_supply_single(),
           root_network = series_resistance(1e3)) |>
  as_tibble() |>
  select(psi_soil, psi_stem, collar, A, gc, g1_eff)
```

No depth profile, and one number instead of a per-layer network. The one layer's
thickness is `leaf_supply_single()`'s only argument, and it is also its
gravitational head — a layer's head is `gravity_head * z_soil_mid` — so the
default of zero lifts water nowhere, which is what a bare leaf wants. It is named
there and nowhere else: passing `soil_depth` to the solve as well is an error
rather than ignored, because two spellings of it could disagree. The resistance is
**per unit leaf area**, because nothing in this model is allowed to scale with
plant size. Raising it makes water harder to get and the leaf closes down:

```{r, fig.height = 3.4}
tibble(resistance = 10^seq(2.5, 4.5, length.out = 25)) |>
  rowwise() |>
  mutate(A = leaf_solve(psi_soil = 1.5, PPFD = 900,
                        supply = leaf_supply_single(),
                        root_network = series_resistance(resistance))$A) |>
  ungroup() |>
  ggplot(aes(resistance, A)) +
  geom_line(linewidth = 0.8, colour = "steelblue4") +
  scale_x_log10() +
  labs(x = expression(soil-to-collar~resistance~(MPa~s~mol^-1~m^2)),
       y = expression(A~(mu*mol~m^-2~s^-1)))
```

This path is also what makes comparison against other optimality models
meaningful: Medlyn, Prentice least-cost and Cowan-Farquhar are all formulated
against a *single* soil potential, so comparing them against a multi-layer root
network would compare two things at once.

## Multiple soil layers

Pass a vector for a layered profile, or a list of vectors for several profiles.
Layers default to 1 m thick with the root carbon split evenly.

```{r}
leaf_solve(psi_soil = list(c(1.0, 1.5, 2.0), c(3.0, 3.5, 4.0)), PPFD = 900) |>
  as_tibble() |>
  select(psi_soil, layers, A, gc, uptake)
```

`psi_soil` in the result is the wettest layer, with `layers` alongside, so the
column stays numeric and the frame stays rectangular.

## The stateful interface

`leaf_solve()` is a wrapper over an R6 object you can drive yourself. This is
what plant does, and it is what you want when you care about the intermediate
state rather than only the operating point.

It is **not** the faster route, which is worth saying because it looks like it
ought to be. Measured over 32 rows, `leaf_solve()` costs 21.5 µs a row and this
loop costs 20.3 — within 6%, because both are dominated by the same thing: each
call across the R boundary costs about 1.1 µs, against 2.8 µs for the whole C++
solve. If you need an inner loop faster than that, what has to go is the number of
R calls per row, not which R function makes them. `?leaf_solve` has the numbers
and the README a section on it.

```{r}
l <- leaf_model()
set_drivers(l, psi_soil = 2.0, PPFD = 900)
l$find_root_collar_psi()

tibble(profit = l$profit_, cost = l$hydraulic_cost_, A = l$assim_colimited_)
```

The object exposes the whole model, not just the operating point. The stem
vulnerability curve is defined for any potential:

```{r, fig.height = 3.4}
tibble(psi = seq(0, 6, length.out = 200)) |>
  mutate(k = vapply(psi, l$proportion_of_conductivity, numeric(1))) |>
  ggplot(aes(psi, k)) +
  geom_line(linewidth = 0.8, colour = "darkgreen") +
  labs(x = expression(psi~(MPa)), y = "proportion of conductivity",
       title = "stem vulnerability")
```

The profit function is not. Water flows from the soil to the leaf, so
`psi_stem` must be **at least** the upstream potential; below it the leaf would
be wetter than the soil, and `transpiration()` and `stom_cond_CO2()` duly return
negative values. Evaluate profit there and you get a number, but it is built on a
negative conductance and means nothing. Plot the domain the solve actually
searches:

```{r, fig.height = 3.4}
psi_upstream <- 2.0

tibble(psi = seq(psi_upstream, 6, length.out = 200)) |>
  mutate(profit = vapply(psi, l$profit_psi_stem_TF, numeric(1),
                         psi_upstream = psi_upstream)) |>
  ggplot(aes(psi, profit)) +
  geom_line(linewidth = 0.8, colour = "darkgreen") +
  geom_vline(xintercept = l$opt_psi_stem_, linetype = 2, colour = "grey40") +
  labs(x = expression(psi[stem]~(MPa)), y = "profit",
       title = "the function being maximised")
```

The dashed line is where the solve landed, and the maximum it sits on is
**flat**. That is worth knowing rather than glossing: the profit *value* is far
better determined than its *location*, which is why `profit` is the only reported
field well conditioned across platforms. It is also why the solve does not search
this curve — an error `dp` in a flat maximum's *value* displaces its *location* by
`sqrt(dp/k)`, so locating the peak by comparing heights costs precision that
solving `dprofit = 0` keeps.

## Where to go next

- The model is a set of self-contained C++ headers that need no R at all, so the
  same solve is available from a C++ program or a Python extension. See the
  README.
- `COMPARISON.md` sets this against `plantecophys`, `tealeaves` and `bigleaf`.
- The leaf takes its resistances through the `root_network` argument on either
  path, so a caller with measured or fitted resistances never has to go through a
  root carbon model to reach a leaf.
