---
title: "Leaf gas exchange from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Leaf gas exchange from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_bw(base_size = 11) +
            theme(panel.grid.minor = element_blank(),
                  strip.background = element_rect(fill = "grey95",
                                                  colour = NA)))
```

## What this model does differently

Most leaf gas-exchange models get stomatal conductance from an empirical
function of assimilation, VPD and CO₂ — Ball-Berry, Medlyn, and their
descendants. This one has no such function. It builds an explicit hydraulic path
from the soil, through the roots and the root collar, up the stem to the leaf,
and then asks: **which stem water potential maximises profit?** Profit is
assimilation minus a hydraulic cost that rises as the xylem loses conductivity.
Conductance is whatever falls out of the answer.

So stomatal closure under drought here is not fitted. It is derived, from two
Weibull vulnerability curves, the soil potential, and the plant's own
conductance.

## One solve

```{r}
leaf_solve(psi_soil = 2.0, PPFD = 900) |>
  as_tibble() |>
  glimpse()
```

Drivers in, operating point out. `psi_soil` is a **positive magnitude in MPa** —
more on that below.

Most of those columns are the usual gas-exchange quantities. Two are less usual:

- **`lambda`** is the marginal cost of water, dA/dE, at the operating point —
  how much carbon the leaf gives up per unit of water it declines to transpire.
  It is the quantity the optimisation equalises.
- **`g1_eff`** re-expresses the solved conductance as the `g1` of the Medlyn
  model, which is a convenient common scale if that is the model you are used
  to. It is computed from the solve rather than supplied, so it varies with
  conditions.

## A drought response

`leaf_solve()` is vectorised over its drivers, so a response curve is one call.

```{r}
drought <- leaf_solve(psi_soil = seq(0.2, 5.5, length.out = 40), PPFD = 900) |>
  as_tibble()

drought |>
  select(psi_soil, A, gc) |>
  pivot_longer(c(A, gc), names_to = "variable") |>
  mutate(variable = recode(variable,
                           A = "assimilation~(mu*mol~m^-2~s^-1)",
                           gc = "stomatal~conductance~(mol~m^-2~s^-1)")) |>
  ggplot(aes(psi_soil, value)) +
  geom_line(linewidth = 0.8, colour = "steelblue4") +
  facet_wrap(~variable, scales = "free_y", labeller = label_parsed) +
  labs(x = "soil water potential (MPa)", y = NULL)
```

Note where both hit exactly zero. That is not a floor imposed by a
minimum-conductance parameter — it is the solve reporting that no operating
point both moves water and stays inside the root's hydraulic limit, so the
answer is shut-down.

## The same sweep on a familiar scale

`g1_eff` across the same drought, over the range where the stomata are open:

```{r, fig.height = 3.4}
drought |>
  filter(gc > 0, is.finite(g1_eff)) |>
  ggplot(aes(psi_soil, g1_eff)) +
  geom_line(linewidth = 0.8, colour = "firebrick") +
  labs(x = "soil water potential (MPa)",
       y = expression(effective~g[1]~(kPa^0.5)))
```

It declines as the soil dries, because it is computed from the solve rather than
being a parameter of it.

## Light and VPD together

Because `leaf_solve()` recycles its drivers elementwise, a factorial design is an
`expand_grid()` and a single call — no loop, and nothing to join back afterwards,
since the drivers come back as columns of the answer.

```{r}
expand_grid(PPFD = seq(50, 1800, length.out = 30),
            atm_vpd = c(0.5, 1, 2, 4)) |>
  with(leaf_solve(psi_soil = 1.0, PPFD = PPFD, atm_vpd = atm_vpd)) |>
  ggplot(aes(PPFD, A, colour = factor(atm_vpd))) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_d(end = 0.9, direction = -1) +
  labs(x = expression(PPFD~(mu*mol~m^-2~s^-1)),
       y = expression(A~(mu*mol~m^-2~s^-1)),
       colour = "VPD (kPa)")
```

The curves saturate at different heights because the hydraulic cost of the
transpiration needed to support a given assimilation rises with VPD.

## Traits and numerical settings are separate

The C++ constructor takes seventeen positional arguments, four of which are
tolerances sitting among the physiology. `leaf_traits()` and `leaf_control()`
split them, so a calibration loop varying traits never has to know which of the
seventeen are not traits.

```{r}
vulnerable <- leaf_traits(stem_b = 2.5, psi_crit = 4.0,
                          root_b = 2.5, root_psi_crit = 4.0)

bind_rows(
  default = leaf_solve(psi_soil = 3.0, PPFD = 900),
  vulnerable = leaf_solve(psi_soil = 3.0, PPFD = 900, traits = vulnerable),
  .id = "traits"
) |>
  select(traits, psi_stem, A, gc, g1_eff)
```

`stem_b`/`stem_c` and `root_b`/`root_c` are **two different vulnerability
curves** — the stem pair drives the hydraulic cost, the root pair drives uptake.
They are named rather than left as an unmarked `b`/`c` because using one where
the other belongs is a mistake that has been made here and reached a manuscript
draft.

Tolerances go the other way, through `leaf_control()`:

```{r}
tibble(
  setting = c("default", "tight"),
  psi_stem = c(
    leaf_solve(psi_soil = 2.0, PPFD = 900)$psi_stem,
    leaf_solve(psi_soil = 2.0, PPFD = 900,
               control = leaf_control(GSS_tol_abs = 1e-5))$psi_stem
  )
)
```

Those are **identical**, and this passage used to say they differed in the fourth
decimal. They did, until the collar solve stopped searching profit and started
solving its own first-order condition instead: the operating point is now
determined to about `1e-12` rather than to `GSS_tol_abs`, which no longer reaches
the production path at all. It survives as the threshold below which the feasible
interval is treated as a single point, and for two single-layer optimisers that
are off this path.

That is not only a precision story — it is what makes the next section possible.
An argmax resolved only to `1e-3` moves in discontinuous jumps as a trait varies,
so its derivative comes back as noise, or worse, smooth and sign-inverted.

## Water potentials are positive magnitudes

There is one representation throughout: every ψ is a positive number of MPa,
including `psi_soil`, `psi_crit` and the collar potential. Where an equation
needs one potential to oppose another, the minus sign is in the equation.

This is asserted rather than documented, which matters if you are porting a
script that passed signed potentials:

```{r, error = TRUE}
leaf_solve(psi_soil = -2.0, PPFD = 900)
```

An error, not a quiet `abs()`. A model that accepted both conventions would let
that script return a plausible wrong number instead.

## A bare leaf: one soil potential, one resistance

Everything above went through the multi-layer root network, which is what plant
uses. The other supply path collapses the whole soil-to-collar journey into a
single resistance — and note that it is driven through **the same `root_network`
argument**, because both paths take their resistances as a per-call driver:

```{r}
leaf_solve(psi_soil = 1.5, PPFD = 900,
           supply = leaf_supply_single(),
           root_network = series_resistance(1e3)) |>
  as_tibble() |>
  select(psi_soil, psi_stem, collar, A, gc, g1_eff)
```

No depth profile, and one number instead of a per-layer network. `soil_depth` is
the only argument that differs between the two paths, and passing it here is an
error rather than ignored: there is no depth profile for anything to read. The
resistance is **per unit leaf area**, because nothing in this model is allowed to
scale with plant size. Raising it makes water harder to get and the leaf closes
down:

```{r, fig.height = 3.4}
tibble(resistance = 10^seq(2.5, 4.5, length.out = 25)) |>
  rowwise() |>
  mutate(A = leaf_solve(psi_soil = 1.5, PPFD = 900,
                        supply = leaf_supply_single(),
                        root_network = series_resistance(resistance))$A) |>
  ungroup() |>
  ggplot(aes(resistance, A)) +
  geom_line(linewidth = 0.8, colour = "steelblue4") +
  scale_x_log10() +
  labs(x = expression(soil-to-collar~resistance~(MPa~s~mol^-1~m^2)),
       y = expression(A~(mu*mol~m^-2~s^-1)))
```

This path is also what makes comparison against other optimality models
meaningful: Medlyn, Prentice least-cost and Cowan-Farquhar are all formulated
against a *single* soil potential, so comparing them against a multi-layer root
network would compare two things at once.

The choice is made when the leaf is built, not afterwards. There is no
`leaf$supply_kind <- "single"`:

```{r, error = TRUE}
l <- leaf_model()
l$supply_kind <- "single"
```

Flipping the tag alone would leave the other path's state configured and silently
ignored — so instead both `leaf_supply_*()` reconfigure the object completely,
and the object can never be in a state where the tag and the supply disagree.

## Multiple soil layers

Pass a vector for a layered profile, or a list of vectors for several profiles.
Layers default to 1 m thick with the root carbon split evenly.

```{r}
leaf_solve(psi_soil = list(c(1.0, 1.5, 2.0), c(3.0, 3.5, 4.0)), PPFD = 900) |>
  as_tibble() |>
  select(psi_soil, layers, A, gc, uptake)
```

`psi_soil` in the result is the wettest layer, with `layers` alongside, so the
column stays numeric and the frame stays rectangular.

## The stateful interface

`leaf_solve()` is a wrapper over an R6 object you can drive yourself. This is
what plant does, and it is what you want when you care about the intermediate
state rather than only the operating point.

It is **not** the faster route, which is worth saying because it looks like it
ought to be. Measured over 32 rows, `leaf_solve()` costs 21.5 µs a row and this
loop costs 20.3 — within 6%, because both are dominated by the same thing: each
call across the R boundary costs about 1.1 µs, against 2.8 µs for the whole C++
solve. If you need an inner loop faster than that, what has to go is the number of
R calls per row, not which R function makes them. `?leaf_solve` has the numbers
and the README a section on it.

```{r}
l <- leaf_model()
set_drivers(l, psi_soil = 2.0, PPFD = 900)
l$find_root_collar_psi()

tibble(profit = l$profit_, cost = l$hydraulic_cost_, A = l$assim_colimited_)
```

The object exposes the whole model, not just the operating point. The stem
vulnerability curve is defined for any potential:

```{r, fig.height = 3.4}
tibble(psi = seq(0, 6, length.out = 200)) |>
  mutate(k = vapply(psi, l$proportion_of_conductivity, numeric(1))) |>
  ggplot(aes(psi, k)) +
  geom_line(linewidth = 0.8, colour = "darkgreen") +
  labs(x = expression(psi~(MPa)), y = "proportion of conductivity",
       title = "stem vulnerability")
```

The profit function is not. Water flows from the soil to the leaf, so
`psi_stem` must be **at least** the upstream potential; below it the leaf would
be wetter than the soil, and `transpiration()` and `stom_cond_CO2()` duly return
negative values. Evaluate profit there and you get a number, but it is built on a
negative conductance and means nothing. Plot the domain the solve actually
searches:

```{r, fig.height = 3.4}
psi_upstream <- 2.0

tibble(psi = seq(psi_upstream, 6, length.out = 200)) |>
  mutate(profit = vapply(psi, l$profit_psi_stem_TF, numeric(1),
                         psi_upstream = psi_upstream)) |>
  ggplot(aes(psi, profit)) +
  geom_line(linewidth = 0.8, colour = "darkgreen") +
  geom_vline(xintercept = l$opt_psi_stem_, linetype = 2, colour = "grey40") +
  labs(x = expression(psi[stem]~(MPa)), y = "profit",
       title = "the function being maximised")
```

The dashed line is where the solve landed, and the maximum it sits on is
**flat**. That is worth knowing rather than glossing: the profit *value* is far
better determined than its *location*, which is why `profit` is the only reported
field well conditioned across platforms. It is also why the solve does not search
this curve — an error `dp` in a flat maximum's *value* displaces its *location* by
`sqrt(dp/k)`, so locating the peak by comparing heights costs precision that
solving `dprofit = 0` keeps.

## Trait gradients, and where they stop working

Because the operating point is now located precisely, it can be *differentiated*.
`leaf_gradient()` returns the derivative of each solved output with respect to
each trait:

```{r}
g <- leaf_gradient(psi_soil = 2.0, PPFD = 900,
                   pars = c("vcmax_25", "jmax_25", "stem_b", "cost_scale_TF24"))
round(g$gradient, 5)
```

### Choosing which gradients to compute

**`pars` selects them, and by default you get all of them.** Omit it and
`leaf_gradient()` differentiates every parameter it can:

```{r}
dim(leaf_gradient(psi_soil = 2.0, PPFD = 900)$gradient)
rownames(leaf_gradient(psi_soil = 2.0, PPFD = 900)$gradient)
```

Fifteen on the multi-layer path — the fourteen traits from `leaf_traits()`, plus
`leaf_specific_conductance_max`, which is not a trait but is fitted like one. The
single-potential path adds `resistance`, making sixteen.

`R_d_25`, dark respiration at 25 °C, is one of those fourteen and behaves like any
other trait. So `dY/dvcmax_25` is a **partial** derivative at fixed respiration: a
fit that wants respiration to follow Vcmax moves both and adds the two columns.

**Selecting fewer genuinely costs less**: the work is one solve plus two cheap
evaluations per parameter, so it scales with what you ask for.

```{r}
#| echo: false
bench <- function(pars) {
  t0 <- Sys.time()
  for (i in 1:20) leaf_gradient(psi_soil = 2.0, PPFD = 900, pars = pars)
  as.numeric(Sys.time() - t0, units = "secs") / 20 * 1000
}
tibble::tibble(
  # Counted, not written down: this label said 16 for a while when the answer was
  # 14, which is exactly the way a number beside a chunk rots (the chunk runs, the
  # prose next to it is never asserted).
  request = c(sprintf("`pars = NULL` (all %d)",
                      nrow(leaf_gradient(psi_soil = 2.0, PPFD = 900)$gradient)),
              "4 parameters", "1 parameter"),
  `ms per call` = round(c(bench(NULL),
                          bench(c("vcmax_25", "jmax_25", "stem_b",
                                  "cost_scale_TF24")),
                          bench("vcmax_25")), 2)
) |>
  knitr::kable()
```

So in a calibration, name only the parameters you are fitting. Asking for all
fourteen when you fit three is several times the work.

⚠️ **The five outputs are not selectable** — every call returns the derivative of
`A`, `gc`, `psi_stem`, `collar` and `profit` with respect to each requested
parameter, because they are read off the same solve and cost nothing extra. `pars`
chooses the rows of that matrix, never the columns.

`profit` is the cheapest of the five. It is what the leaf maximises — carbon gained
minus the hydraulic cost of the water that bought it — so it is the objective
rather than an output read at the optimum, and at an interior optimum
`dprofit/dψ = 0` makes its trait derivative the partial `∂profit/∂θ` at fixed ψ.
That is the envelope theorem: the profit row needs no sensitivity of the optimiser,
where the other four do.

Naming something undifferentiable is an error rather than a silent omission, and
it tells you what is available:

```{r, error = TRUE}
leaf_gradient(psi_soil = 2.0, PPFD = 900, pars = "resistance")
```

These are not finite differences of the solve. The outputs are read at the
profit-maximising collar potential, so a trait moves them twice over — directly,
and by moving the optimum itself. Differentiating the optimality condition
`dprofit/dψ = 0` rather than the solved output captures both terms, and for
`cost_scale_TF24`, `beta2`, `stem_b` and `stem_c` the second one is **the entire
answer**: change them and assimilation moves *only* because the leaf chooses a
different operating point.

The interesting part is where the derivation stops being valid. It assumes the
optimum is interior, and in dry soil it often is not — the leaf ends up pinned
against the edge of the range of collar potentials it can physically occupy.
There `dprofit/dψ` is not zero at the answer, the formula is not merely imprecise
but wrong by orders of magnitude, and it does not fail loudly. So the assumption
is tested at every point, and `leaf_gradient()` differences the solve instead
where it does not hold:

```{r, fig.height = 3.4}
sweep <- lapply(seq(1.0, 5.0, by = 0.125), function(p) {
  g <- leaf_gradient(psi_soil = p, PPFD = 900, atm_vpd = 3.0, pars = "stem_b")
  tibble(psi_soil = p, dA_dstem_b = g$gradient["stem_b", "A"],
         method = g$method, status = g$status)
}) |> bind_rows()

ggplot(sweep, aes(psi_soil, dA_dstem_b)) +
  geom_line(colour = "grey60") +
  geom_point(aes(colour = status), size = 2) +
  labs(x = expression(psi[soil]~(MPa)),
       y = expression(dA/d*italic(stem_b)),
       colour = NULL,
       title = "sensitivity to stem vulnerability, and where the optimum pins")
```

Left to right: in wet soil a less vulnerable stem is worth about two units of
assimilation per MPa of `stem_b`. The sensitivity rises as the soil dries, because
vulnerability matters more when water is scarce, peaks near 2.9 MPa, eases off —
and then falls off a cliff. Past about 3.9 MPa the optimum is **pinned**: the leaf
is stuck at the wettest collar potential it can physically reach, and a stem trait
it has no freedom to exploit stops mattering at all.

Two honest notes about that flat tail. The true gradient there is smaller than the
`5e-05` the plotted points reach, so what is drawn is the finite difference's own
noise about zero — which is why it wobbles in sign. Treat it as "zero", not as a
small effect. And the
transition itself is a genuine **kink** in trait space — a real
non-differentiable set that belongs to the constrained problem rather than to this
implementation. A gradient-based fit that wanders across it is optimising through
a corner, and no amount of care in the derivative changes that. What
`leaf_gradient()` can do is tell you when you are there, which the naive finite
difference cannot: `status` marks these points, and `g$stationarity` reports the
measured distance from stationarity the classification rests on.

## Where to go next

- The model is a set of self-contained C++ headers that need no R at all, so the
  same solve is available from a C++ program or a Python extension. See the
  README.
- `COMPARISON.md` sets this against `plantecophys`, `tealeaves` and `bigleaf`.
- Both supply paths now take resistances, through the same `root_network`
  argument, so a caller with measured or fitted resistances never has to go
  through a root carbon model to reach a leaf. The one asymmetry left is
  `gravity_head`, which `leaf_supply_single()` still takes as configuration
  because this path has no depth profile to derive it from — see its
  documentation.
