---
title: "Leaf gas exchange from R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Leaf gas exchange from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, dpi = 120, fig.align = "center")
```

```{r setup, message = FALSE}
library(phylloptim)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_bw(base_size = 11) +
            theme(panel.grid.minor = element_blank(),
                  strip.background = element_rect(fill = "grey95",
                                                  colour = NA)))
```

## What this model does differently

Most leaf gas-exchange models obtain stomatal conductance from an empirical
function of assimilation, VPD and CO₂. Ball-Berry, Medlyn and their descendants
all work that way. This model has no such function.

Instead it builds an explicit hydraulic path from the soil, through the roots and
the root collar, up the stem to the leaf. Then it asks which stem water potential
maximises profit, where profit is assimilation minus a hydraulic cost that rises
as the xylem loses conductivity. Conductance is whatever falls out of the answer.

Stomatal closure under drought is therefore not fitted here. It is derived, from
two Weibull vulnerability curves, the soil potential, and the plant's own
conductance.

## What is on this page

| Section | What it covers |
|---|---|
| *One solve* to *Light and VPD together* | driving the model and reading the result |
| *Traits and numerical settings are separate* | which arguments are physiology and which are numerics |
| *Water potentials are positive magnitudes* | the one convention that will break a ported script |
| *A bare leaf* | the simpler supply path, for a leaf with no root profile |
| *Trait gradients* | exact derivatives, and when they are worth using |
| *A faster method* | the closed form: what it costs in accuracy, and when it refuses |

`vignette("the-models")` is the reference for what is being solved.
`vignette("fitting")` works through a calibration.

## One solve

```{r}
leaf_solve(psi_soil = 2.0, PPFD = 900) |>
  as_tibble() |>
  glimpse()
```

Drivers in, operating point out. `psi_soil` is a positive magnitude in MPa, which
a later section covers.

Most of those columns are the usual gas-exchange quantities. Two are less usual.

`lambda` is the marginal cost of water, dA/dE, at the operating point. It is how
much carbon the leaf gives up per unit of water it declines to transpire, and it
is the quantity the optimisation equalises.

`g1_eff` re-expresses the solved conductance as the `g1` of the Medlyn model,
which is a convenient common scale if that is the model you are used to. It is
computed from the solve rather than supplied, so it varies with conditions.

## A drought response

`leaf_solve()` is vectorised over its drivers, so a response curve is one call.

```{r}
drought <- leaf_solve(psi_soil = seq(0.2, 5.5, length.out = 40), PPFD = 900) |>
  as_tibble()

drought |>
  select(psi_soil, A, gc) |>
  pivot_longer(c(A, gc), names_to = "variable") |>
  mutate(variable = recode(variable,
                           A = "assimilation~(mu*mol~m^-2~s^-1)",
                           gc = "stomatal~conductance~(mol~m^-2~s^-1)")) |>
  ggplot(aes(psi_soil, value)) +
  geom_line(linewidth = 0.8, colour = "steelblue4") +
  facet_wrap(~variable, scales = "free_y", labeller = label_parsed) +
  labs(x = "soil water potential (MPa)", y = NULL)
```

Note where both hit exactly zero. That is not a floor imposed by a
minimum-conductance parameter. It is the solve reporting that no operating point
both moves water and stays inside the root's hydraulic limit, so the answer is
shut-down.

## The same sweep on a familiar scale

`g1_eff` across the same drought, over the range where the stomata are open:

```{r, fig.height = 3.4}
drought |>
  filter(gc > 0, is.finite(g1_eff)) |>
  ggplot(aes(psi_soil, g1_eff)) +
  geom_line(linewidth = 0.8, colour = "firebrick") +
  labs(x = "soil water potential (MPa)",
       y = expression(effective~g[1]~(kPa^0.5)))
```

It declines as the soil dries, because it is computed from the solve rather than
being a parameter of it.

## Light and VPD together

Because `leaf_solve()` recycles its drivers elementwise, a factorial design is an
`expand_grid()` and a single call. There is no loop, and nothing to join back
afterwards, because the drivers come back as columns of the answer.

```{r}
expand_grid(PPFD = seq(50, 1800, length.out = 30),
            atm_vpd = c(0.5, 1, 2, 4)) |>
  with(leaf_solve(psi_soil = 1.0, PPFD = PPFD, atm_vpd = atm_vpd)) |>
  ggplot(aes(PPFD, A, colour = factor(atm_vpd))) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_d(end = 0.9, direction = -1) +
  labs(x = expression(PPFD~(mu*mol~m^-2~s^-1)),
       y = expression(A~(mu*mol~m^-2~s^-1)),
       colour = "VPD (kPa)")
```

The curves saturate at different heights because the hydraulic cost of the
transpiration needed to support a given assimilation rises with VPD.

## Traits and numerical settings are separate

The C++ constructor takes fifteen positional arguments, four of which are
tolerances sitting among the physiology. There are also four traits it does not
take at all, which have to be assigned afterwards. `leaf_traits()` and
`leaf_control()` split the first fifteen and supply the missing four, so a
calibration loop varying traits never has to know either fact.

```{r}
vulnerable <- leaf_traits(stem_P50 = 2.2, root_P50 = 2.2)

bind_rows(
  default = leaf_solve(psi_soil = 3.0, PPFD = 900),
  vulnerable = leaf_solve(psi_soil = 3.0, PPFD = 900, traits = vulnerable),
  .id = "traits"
) |>
  select(traits, psi_stem, A, gc, g1_eff)
```

`stem_P50`/`stem_c` and `root_P50`/`root_c` are two different vulnerability
curves. The stem pair drives the hydraulic cost; the root pair drives uptake.
They are named explicitly rather than left as an unmarked `b`/`c` pair, because
using one where the other belongs is easy to do and produces a plausible wrong
answer rather than an error.

Tolerances go the other way, through `leaf_control()`:

```{r}
tibble(
  setting = c("default", "tight"),
  psi_stem = c(
    leaf_solve(psi_soil = 2.0, PPFD = 900)$psi_stem,
    leaf_solve(psi_soil = 2.0, PPFD = 900,
               control = leaf_control(GSS_tol_abs = 1e-5))$psi_stem
  )
)
```

Those two are identical, because the collar solve solves its own first-order
condition rather than searching profit. The operating point is determined to about
1e-12, so `GSS_tol_abs` does not reach the production path at all.

It is read in two places, both on the collar route: the threshold below which the
feasible collar interval is treated as a single point, and the tolerance of a
fallback search used when neither bracket endpoint has a usable gradient. It does
not reach the stem route, which root-finds its own first-order condition at a
hardcoded tolerance, so tightening it to sharpen a stem-route argmax does
nothing.

This is not only a precision story. It is what makes the trait gradients below
possible. An argmax resolved only to 1e-3 moves in discontinuous jumps as a trait
varies, so its derivative comes back as noise, or worse, as something smooth,
plausible and sign-inverted.

## Water potentials are positive magnitudes

There is one representation throughout: every ψ is a positive number of MPa,
including `psi_soil`, `psi_crit` and the collar potential. Where an equation
needs one potential to oppose another, the minus sign is in the equation.

This is asserted rather than documented, which matters if you are porting a
script that passed signed potentials:

```{r, error = TRUE}
leaf_solve(psi_soil = -2.0, PPFD = 900)
```

That is an error rather than a quiet `abs()`. A model accepting both conventions
would let such a script return a plausible wrong number instead.

## A bare leaf: one soil potential, one resistance

Everything above went through the multi-layer root network, which is what `plant`
uses. The other supply path collapses the whole soil-to-collar journey into a
single resistance. It is driven through the same `root_network` argument, because
both paths take their resistances as a per-call driver:

```{r}
leaf_solve(psi_soil = 1.5, PPFD = 900,
           supply = leaf_supply_singlelayer(),
           root_network = series_resistance(1e3)) |>
  as_tibble() |>
  select(psi_soil, psi_stem, collar, A, gc, g1_eff)
```

No depth profile, and one number instead of a per-layer network. `soil_depth` is
the only argument that differs between the two paths, and passing it here is an
error rather than ignored, because there is no depth profile for anything to read.

The resistance is per unit leaf area, because nothing in this model is allowed to
scale with plant size. Raising it makes water harder to get, and the leaf closes
down:

```{r, fig.height = 3.4}
tibble(resistance = 10^seq(2.5, 4.5, length.out = 25)) |>
  rowwise() |>
  mutate(A = leaf_solve(psi_soil = 1.5, PPFD = 900,
                        supply = leaf_supply_singlelayer(),
                        root_network = series_resistance(resistance))$A) |>
  ungroup() |>
  ggplot(aes(resistance, A)) +
  geom_line(linewidth = 0.8, colour = "steelblue4") +
  scale_x_log10() +
  labs(x = expression(soil-to-collar~resistance~(MPa~s~mol^-1~m^2)),
       y = expression(A~(mu*mol~m^-2~s^-1)))
```

This path is also what makes comparison against other optimality models
meaningful. Medlyn, Prentice least-cost and Cowan-Farquhar are all formulated
against a single soil potential, so comparing them against a multi-layer root
network would compare two things at once.

The choice is made when the leaf is built rather than afterwards. There is no
settable `leaf$supply_kind`:

```{r, error = TRUE}
l <- leaf_model()
l$supply_kind <- "single"
```

Flipping the tag alone would leave the other path's state configured and silently
ignored. Both `leaf_supply_*()` functions therefore reconfigure the object
completely, so it can never be in a state where the tag and the supply disagree.

## Multiple soil layers

Pass a vector for a layered profile, or a list of vectors for several profiles.
Layers default to 1 m thick with the root carbon split evenly.

```{r}
leaf_solve(psi_soil = list(c(1.0, 1.5, 2.0), c(3.0, 3.5, 4.0)), PPFD = 900) |>
  as_tibble() |>
  select(psi_soil, layers, A, gc, uptake)
```

`psi_soil` in the result is the wettest layer, with `layers` alongside, so the
column stays numeric and the frame stays rectangular.

## The stateful interface

`leaf_solve()` is a wrapper over an R6 object you can drive yourself. This is what
`plant` does, and it is what you want when you care about the intermediate state
rather than only the operating point.

It is not the faster route, which is worth saying because it looks as though it
ought to be. Measured over 32 rows, `leaf_solve()` costs 21.5 µs a row and this
loop costs 20.3, so they are within 6% of each other. Both are dominated by the
same thing: a call across the R boundary costs about 1.1 µs, against 2.8 µs for
the whole C++ solve. If you need an inner loop faster than that, what has to go is
the number of R calls per row rather than which R function makes them.
`?leaf_solve` has the numbers, and the README has a section on it.

```{r}
l <- leaf_model()
set_drivers(l, psi_soil = 2.0, PPFD = 900)
l$optimise()          # or l$find_root_collar_psi(), which plant's sources spell

tibble(profit = l$profit_, cost = l$hydraulic_cost_, A = l$assim_colimited_)
```

The object exposes the whole model, not just the operating point. The stem
vulnerability curve is defined for any potential:

```{r, fig.height = 3.4}
tibble(psi = seq(0, 6, length.out = 200)) |>
  mutate(k = vapply(psi, l$proportion_of_conductivity, numeric(1))) |>
  ggplot(aes(psi, k)) +
  geom_line(linewidth = 0.8, colour = "darkgreen") +
  labs(x = expression(psi~(MPa)), y = "proportion of conductivity",
       title = "stem vulnerability")
```

The profit function is not defined everywhere. Water flows from the soil to the
leaf, so `psi_stem` must be at least the upstream potential. Below it the leaf
would be wetter than the soil, and `transpiration()` and `stom_cond_CO2()` duly
return negative values. Evaluating profit there returns a number, but the number
is built on a negative conductance and means nothing. Plot the domain the solve
actually searches:

```{r, fig.height = 3.4}
psi_upstream <- 2.0

tibble(psi = seq(psi_upstream, 6, length.out = 200)) |>
  mutate(profit = vapply(psi, l$profit_psi_stem_TF, numeric(1),
                         psi_upstream = psi_upstream)) |>
  ggplot(aes(psi, profit)) +
  geom_line(linewidth = 0.8, colour = "darkgreen") +
  geom_vline(xintercept = l$opt_psi_stem_, linetype = 2, colour = "grey40") +
  labs(x = expression(psi[stem]~(MPa)), y = "profit",
       title = "the function being maximised")
```

The dashed line is where the solve landed, and the maximum it sits on is flat.
That flatness has a practical consequence: the profit *value* is much better
determined than its *location*. An error `dp` in the value displaces the location
by `sqrt(dp/k)` for curvature `k`, so locating the peak by comparing heights costs
precision that solving `dprofit = 0` keeps. It is why the solve does not search
this curve. `vignette("the-models")` Appendix C has the measurements.

## Trait gradients, and where they stop working

Because the operating point is now located precisely, it can be *differentiated*.
`leaf_gradient()` returns the derivative of each solved output with respect to
each trait:

```{r}
g <- leaf_gradient(psi_soil = 2.0, PPFD = 900,
                   pars = c("vcmax_25", "jmax_25", "stem_P50", "TF24_cost_scale"))
round(g$gradient, 5)
```

### Choosing which gradients to compute

`pars` selects them, and by default you get all of them. Omit it and
`leaf_gradient()` differentiates every parameter it can:

```{r}
dim(leaf_gradient(psi_soil = 2.0, PPFD = 900)$gradient)
rownames(leaf_gradient(psi_soil = 2.0, PPFD = 900)$gradient)
```

That is `r length(leaf_traits())` traits from `leaf_traits()`, plus
`leaf_specific_conductance_max`, which is not a trait but is fitted like one,
giving `r length(leaf_traits()) + 1L` on the multi-layer path. The
single-potential path adds `resistance`, and `model = "CF77"` adds
`CF77_lambda_`, that curve's prescribed marginal value of water. These counts are
computed rather than written out, for the reason the hidden chunk below records.

`R_d_25`, dark respiration at 25 °C, is one of the traits and behaves like any
other. So `dY/dvcmax_25` is a partial derivative at fixed respiration. A fit that
wants respiration to follow Vcmax moves both and adds the two columns.

Selecting fewer parameters genuinely costs less. The work is one solve plus two
cheap evaluations per parameter, so it scales with what you ask for.

```{r}
#| echo: false
bench <- function(pars) {
  t0 <- Sys.time()
  for (i in 1:20) leaf_gradient(psi_soil = 2.0, PPFD = 900, pars = pars)
  as.numeric(Sys.time() - t0, units = "secs") / 20 * 1000
}
tibble::tibble(
  # Counted, not written down: this label said 16 for a while when the answer was
  # 14, and the prose above it went on saying 14 after the answer became 16. That
  # is exactly the way a number beside a chunk rots -- the chunk runs, the prose
  # next to it is never asserted -- so the prose uses inline R now too.
  request = c(sprintf("`pars = NULL` (all %d)",
                      nrow(leaf_gradient(psi_soil = 2.0, PPFD = 900)$gradient)),
              "4 parameters", "1 parameter"),
  `ms per call` = round(c(bench(NULL),
                          bench(c("vcmax_25", "jmax_25", "stem_P50",
                                  "TF24_cost_scale")),
                          bench("vcmax_25")), 2)
) |>
  knitr::kable()
```

So in a calibration, name only the parameters you are fitting. Asking for all of
them when you fit three is several times the work.

The five outputs are not selectable. Every call returns the derivative of `A`,
`gc`, `psi_stem`, `collar` and `profit` with respect to each requested parameter,
because they are read off the same solve and cost nothing extra. `pars` chooses
the rows of that matrix, never the columns.

The first four are what a gas-exchange calibration observes. `profit` is the
objective, and it is there for a demographic consumer: `plant` bills carbon from
the leaf's `profit` rather than its assimilation. It is also the one output the
envelope theorem reaches, so at an interior optimum its gradient is the direct
partial alone. `?leaf_gradient` explains that, along with the measurement that
makes it worth doing rather than merely true.

### A collar potential you supply

If your model carries the collar potential rather than solving for it each step,
pass it and the gradient is taken there:

```{r}
g <- leaf_gradient(psi_soil = 2.0, PPFD = 900, pars = "vcmax_25", psi = 2.7)
c(status = g$status, method = g$method)
```

`psi` is exogenous, so the answer is the partial derivative at fixed collar, plus
`dpsi_dtheta` if the collar you imposed itself moves with the traits.
`stationarity` still comes back, and on this path it says how far the point you
chose sits from the optimum. `M`, `H` and `dY_dpsi` come back too, so a caller
integrating its own sensitivity of `psi` has the coefficients.

Hand back the collar the solver found, with the response it derived, and the two
paths agree exactly:

```{r}
a <- leaf_gradient(psi_soil = 2.0, PPFD = 900, pars = "vcmax_25")
b <- leaf_gradient(psi_soil = 2.0, PPFD = 900, pars = "vcmax_25",
                   psi = a$psi, dpsi_dtheta = -a$M / a$H)
identical(a$gradient, b$gradient)
```

A `psi` outside the feasible collar interval gets no gradient. It is clamped, so
the collar actually used moves with the *bound* rather than with your
`dpsi_dtheta`, and the direct term alone would be plausible and wrong. `status`
reports `"clamped"`, the gradient is `NA`, and `g$psi` tells you where the collar
was pulled to.

Naming something that cannot be differentiated is an error rather than a silent
omission, and the message says what is available instead:

```{r, error = TRUE}
leaf_gradient(psi_soil = 2.0, PPFD = 900, pars = "resistance")
```

These are not finite differences of the solve. The outputs are read at the
profit-maximising collar potential, so a trait moves them twice over: directly,
and by moving the optimum itself. Differentiating the optimality condition
`dprofit/dψ = 0`, rather than the solved output, captures both terms.

For `TF24_cost_scale`, `TF24_beta2`, `stem_P50` and `stem_c` the second term is
the entire answer. Change any of them and assimilation moves only because the leaf
chooses a different operating point.

The interesting part is where the derivation stops being valid. It assumes the
optimum is interior, and in dry soil it often is not: the leaf ends up at the very
edge of the range of collar potentials it can physically occupy. There
`dprofit/dψ` is not zero at the answer, and the formula is not merely imprecise
but wrong by orders of magnitude, without failing loudly. The assumption is
therefore tested at every point, and `leaf_gradient()` differences the solve
instead where it does not hold:

```{r, fig.height = 3.4}
sweep <- lapply(seq(1.0, 5.0, by = 0.125), function(p) {
  g <- leaf_gradient(psi_soil = p, PPFD = 900, atm_vpd = 3.0, pars = "stem_P50")
  tibble(psi_soil = p, dA_dstem_P50 = g$gradient["stem_P50", "A"],
         method = g$method, status = g$status)
}) |> bind_rows()

ggplot(sweep, aes(psi_soil, dA_dstem_P50)) +
  geom_line(colour = "grey60") +
  geom_point(aes(colour = status), size = 2) +
  labs(x = expression(psi[soil]~(MPa)),
       y = expression(dA/d*italic(stem_P50)),
       colour = NULL,
       title = "sensitivity to stem vulnerability, and where the optimum pins")
```

Read that left to right. In wet soil, a less vulnerable stem is worth about two
units of assimilation per MPa of `stem_P50`. The sensitivity rises as the soil
dries, because vulnerability matters more when water is scarce, peaks near 2.9 MPa,
eases off, and then falls away sharply. Past about 3.9 MPa the optimum is a bound:
the leaf is stuck at the wettest collar potential it can physically reach, and a
stem trait it has no freedom to exploit stops mattering at all.

Two things about that flat tail are worth stating plainly. The true gradient there
is smaller than the 5e-05 the plotted points reach, so what is drawn is the finite
difference's own noise about zero, which is why it wobbles in sign. Read it as
zero rather than as a small effect.

The transition itself is a genuine kink in trait space, and a real
non-differentiable set. It belongs to the constrained problem rather than to this
implementation. A gradient-based fit that wanders across it is optimising through
a corner, and no amount of care in the derivative changes that. What
`leaf_gradient()` can do is tell you when you are there, which a naive finite
difference cannot: `status` marks these points, and `g$stationarity` reports the
measured distance from stationarity that the classification rests on.

## A faster method: the closed form

Everything above reaches the operating point the same way: a root-find of the
first-order condition over the feasible interval. That is the `method = "exact"`
path, and it is the default. `set_model()` takes a third argument, and
`"closed"` selects a different route to the same optimum.

The idea is the one `vignette("the-models")` builds the whole comparison on. Every
model here satisfies `dA/dE = lambda` at its optimum, and *given* `lambda` the
answer collapses to the Medlyn relation `ci/ca = xi/(xi + sqrt(D))` with
`xi = sqrt(Q/lambda)`. The difficulty is that `lambda` depends on the `psi` you are
solving for — but where it is a power law in `psi` at the wet end, that
circularity inverts explicitly. One Newton step on the supply-minus-demand
residual finishes the job, and a search over the bracket is not needed at all.

```{r}
cf <- leaf_model(supply = leaf_supply_singlelayer())
set_drivers(cf, psi_soil = 0.5, PPFD = 1500, atm_vpd = 1,
            root_network = series_resistance(1500))

cf$set_model("TF24", "stem", "closed")
cf$optimise()
tibble(method = cf$model_method(), psi_stem = cf$opt_psi_stem_,
       A = cf$assim_colimited_, fell_back = cf$last_solve_fell_back_)
```

### When it applies, and when it refuses

It is refused rather than approximated wherever it does not hold, and the
refusals carry their reasons:

- **`TF24`, `CF77` and `TF24_floor` only, of the eight cost curves.** The three-way
  split is worth knowing, because only one part of it is a gap in the code. A
  closed form exists exactly when `h'(A)`, the derivative of the benefit link, does
  not depend on the solution. For `JS22`, `CMax` and `ProfitMax` it does not — so
  the inversion *exists* for them and is simply not written yet. For `SOX` and
  `JW26` it does: their link is `log`, so `h'(A) = 1/A`, and `lambda` therefore
  carries the assimilation the solve is for. Given `lambda -> xi -> ci -> A` the
  dependency closes into a fixed point, which can be iterated but not inverted.
  Those two **cannot** have a closed form.

  ⚠️ **`TF24_floor` needs `$TF24_floor_lambda_o > 0`, and its tail is worse than
  the other two's.** Its wet-end `lambda` is `lambda_o` — a constant, so the same
  `n = 0` case as `CF77` — and at `lambda_o = 0` that limit is zero, `xi =
  sqrt(Q/lambda)` diverges, and there is nothing to expand about; that case is
  `TF24`, which has its own start, and the refusal points there. Where it does
  apply, the leading order discards the whole `Θ̃'(ψ)/K(ψ)` term, which at the
  DEFAULT `TF24_cost_scale` dominates `lambda_o` at the operating point. Measured
  over a 150-row driver grid, the worst served error is 2.9e-01 at
  `TF24_cost_scale = 0.5` (against `CF77`'s own 2.4e-01) and **5.2e-01** at the
  default 7.5. The `ci/ca` guard does not catch that, because it reads the closed
  form's own `lambda` and stays plausible while `psi*` is wrong — so on this curve
  the fast arm is a measurement tool rather than a production path.
- **The stem route only, and one soil layer only.** The inversion is against a
  supply measured from `psi_soil`; it says nothing about the soil-to-collar path,
  and the multi-layer closed form does not exist.
- **Not with the energy balance on.** Fick's law divides by the leaf-to-air
  deficit, which depends on leaf temperature, which depends on transpiration —
  which is what the inversion solves for. So `D` is not known at the moment of
  inversion and the form is not closed. This is a statement about the form rather
  than about its accuracy: measured with the refusal bypassed, the error reaches
  2.3 MPa in `psi_stem` even in the wettest bin, against 0.04 MPa on the
  prescribed-temperature path.

```{r, error = TRUE}
cf$set_model("SOX", "stem", "closed")
cf$optimise()
```

### What it costs, and what it buys

The guard on validity tests an **output** — `ci/ca` at the answer — so it can only
be applied after solving, and a rejected row falls back to the exact solve. That
makes the fallback fraction `phi` part of the price: the realised speedup is
`1/[phi + (1 - phi)/s]` rather than the ceiling `s`.

Measured in `tests/cpp/bench_solve`, with both arms in one process on a 96-point
single-layer driver grid:

| curve | exact | closed | realised | `phi` | ceiling at `phi = 0` |
|---|---|---|---|---|---|
| `TF24` | 2.65 µs | 1.16 µs | **2.3×** | 0.27 | 4.4× |
| `CF77` | 2.74 µs | 1.37 µs | **2.0×** | 0.39 | 5.2× |

Two-fold, in other words — not the order of magnitude an earlier version of this
code advertised. Those figures predated the current solver, which root-finds the
first-order condition where it used to search the objective, and they also priced
the inversion alone rather than a solve a caller can read outputs from.

`phi` is reported per leaf, so you can measure it on your own drivers rather than
taking the grid's:

```{r}
sweep_leaf <- leaf_model(supply = leaf_supply_singlelayer())
sweep_leaf$set_model("TF24", "stem", "closed")
for (p in seq(0.25, 5, length.out = 20)) {
  set_drivers(sweep_leaf, psi_soil = p, PPFD = 1500, atm_vpd = 1,
              root_network = series_resistance(1500))
  sweep_leaf$optimise()
}
tibble(calls = sweep_leaf$closed_form_calls_,
       fell_back = sweep_leaf$closed_form_fallbacks_,
       phi = sweep_leaf$closed_form_fallback_fraction())
```

### Accuracy, and one failure that is not a tolerance

Three approximations sit inside the closed form: the power law for `lambda`, the
single Newton step, and the Medlyn collapse itself — whose `3*Gstar` group is the
electron-transport-limited `dA/dci`, so it is not exact under colimitation. The
last of those is the binding one, and the error is organised by `ci/ca`:

```{r}
grid <- expand.grid(psi_soil = c(0.25, 0.5, 1, 2, 3, 4, 5),
                    PPFD = c(100, 500, 1500),
                    atm_vpd = c(0.5, 1, 2, 4))

solve_both <- function(psi_soil, PPFD, atm_vpd) {
  one <- function(method) {
    l <- leaf_model(supply = leaf_supply_singlelayer())
    set_drivers(l, psi_soil = psi_soil, PPFD = PPFD, atm_vpd = atm_vpd,
                root_network = series_resistance(1500))
    l$set_model("TF24", "stem", method)
    l$optimise()
    l
  }
  ex <- one("exact")
  cl <- one("closed")
  tibble(ci_ca = ex$ci_ / ex$ca_, fell_back = cl$last_solve_fell_back_,
         err_A = abs(cl$assim_colimited_ - ex$assim_colimited_) /
           max(1, abs(ex$assim_colimited_)),
         err_psi = abs(cl$opt_psi_stem_ - ex$opt_psi_stem_))
}

res <- Map(solve_both, grid$psi_soil, grid$PPFD, grid$atm_vpd) |>
  bind_rows() |>
  filter(is.finite(ci_ca))

res |>
  mutate(bin = cut(ci_ca, c(0, 0.25, 0.4, 0.5, 0.6, 0.8, 1))) |>
  group_by(bin) |>
  summarise(n = n(), fell_back = sum(fell_back),
            mean_err_A = mean(err_A), worst_err_A = max(err_A),
            worst_err_psi = max(err_psi), .groups = "drop")
```

Read it as three regimes. Above `ci/ca` about 0.6 the closed form agrees with the
exact solve to well under a percent in assimilation on average, and a few percent
at worst. Between 0.5 and 0.6 the mean is a couple of percent. And below about 0.4
the guard has fired on nearly every row, so the tabulated error collapses to the
solver's own floor — those rows *are* the exact solve.

⚠️ **The failure below `ci/ca` about 0.3 is topological, not a matter of
tolerance.** There the exact optimum turns back toward shutdown while the
inversion runs on to `psi_crit`: the fixed point the closed form is looking for
has ceased to exist, so the answer is wrong rather than imprecise, and no amount
of refinement recovers it. That is why there is a guard and a fallback at all
rather than a tighter convergence criterion — and it is why `$last_solve_fell_back_`
is worth reading if you are going to trust an individual row.

### The caveat that would stop you using this in a demographic model

`plant` differentiates the operating point with respect to plant state, and that
needs the argmax to vary *smoothly* with its inputs. Sweeping `kmax` — which is
how height reaches the leaf — over ±30% and taking the mean absolute second
difference of `opt_psi_stem_`:

| the sweep | exact | closed |
|---|---|---|
| never crosses the guard | 6.1e-06 | 7.2e-06 |
| crosses the guard | 1.0e-05 | 1.0e-03 |

The closed form's own arithmetic is barely rougher than the solve, at about 1.2×.
What costs two orders of magnitude is the **fallback boundary**: the guard tests an
output, so the set of drivers on which the method switches is a surface in driver
space, and the argmax jumps as a smooth input carries the leaf across it. A
mixture of two methods is discontinuous even where both are individually smooth.

So this is usable for a large batch of independent operating points, and it is
**not** usable underneath a derivative. `leaf_gradient()` reflects that: seating a
model resets the method to `"exact"`, so a gradient always differentiates the
exact solve whatever was seated before.

## What the model assumes

Every model simplifies. The assumptions most worth keeping in mind when reading
output from this one are:

- **Stomatal behaviour is optimal and instantaneous.** The leaf sits at the
  maximum of its objective at every point, with no cost of adjustment and no lag.
- **The leaf is intensive.** Every input is per unit leaf area, or is a
  dimensionless driver. Nothing scales with plant size, so whole-plant allometry
  belongs to the caller.
- **The hydraulic path is two Weibull curves in steady state.** There is no
  capacitance and no refilling.
- **The default cost curve is one hypothesis among several.** `TF24` is the
  default because it is what `plant` runs, not because it has been shown to be
  the right account of what a leaf is spending. `vignette("the-models")` has the
  other six and how they differ.
- **The default traits are *Eucalyptus saligna*.** They are a starting point
  rather than a recommendation, and the same is true of both default supply
  networks.

## Where to go next

- **I want to know what is being solved.** `vignette("the-models")` is the
  reference: the objective, the eight cost curves, the first-order conditions,
  and the two optional gates.
- **I want to fit this to data.** `vignette("fitting")` works through a
  calibration, and covers when the exact gradient earns its complexity.
- **I want to call it without R.** The model is a set of self-contained C++
  headers, so the same solve is available from a C++ program or a Python
  extension. `vignette("cpp-interface")` covers it.
- **I want to know how this differs from what I already use.**
  `COMPARISON.md` sets it against `plantecophys`, `tealeaves` and `bigleaf`.

One loose end worth flagging if you are supplying your own hydraulics. Both
supply paths take resistances through the same `root_network` argument, so a
caller with measured or fitted resistances never has to go through a root carbon
model. The single exception is `gravity_head`, which `leaf_supply_singlelayer()`
still takes as configuration, because that path has no depth profile to derive it
from.
